---
title: "College Football Player Size Exploration"
author: "Sean Osier (smo2152)"
date: "2023-04-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(glue)
```

## Load Data

```{r}
# Load Data
players <- read_csv("cfb_players_2022.csv")
records <- read_csv("cfb_team_records_2022.csv")

# Preview Data
players
records
```

## Helper Functions

```{r}

clean_position_data <- function(df) {
  df %>%
    # Clean up position data:
    filter(
      # Exclude "Athletes" (ATH) who don't have a set / decided position:
      position != "ATH",
      # Only 1 player is listed as a designated Punt Returner (PR):
      position != "PR",
    ) %>% 
    mutate(
      position = case_when(
        # Most teams don't list specific Offensive Line (OL) positions:
        position %in% c("C", "G", "OT") ~ "OL",
        # More teams list specific Defensive Line (DL) positions, but not all:
        position %in% c("DE", "DT", "NT") ~ "DL",
        # It's more common for teams to list specific Defensive Back (DB)
        # positions, but again many teams do not:
        position %in% c("CB", "S") ~ "DB",
        # The Full Back (FB) position is antiquated in many modern offenses.
        # For teams that do use the position usage varies and either Running
        # Back (RB) or Tight End (TE) could be appropriate. In triple option
        # offense teams like Air Force and Navy, RB is the closest position. In
        # others TE is closest:
        position %in% c("FB") & team %in% c("Air Force", "Navy") ~ "RB",
        position %in% c("FB") ~ "TE",
        TRUE ~ position
      )
    )
}

calculate_bmi <- function(players) {
  players %>%
    mutate(bmi = 703 * weight / height^2)
}

calculate_summary_metrics <- function(grouped_df) {
  grouped_df %>%
    summarize(
      height = mean(height),
      weight = mean(weight),
      bmi = mean(bmi),
      n = n()
    )
}

calculate_summary_by <- function(players, ...) {
  players %>%
    group_by(...) %>%
    calculate_summary_metrics()
}
```

## Clean & Summarize Data

```{r}
players <- players %>%
  clean_position_data() %>%
  calculate_bmi()

position_summary <- players %>%
  calculate_summary_by(position)

team_summary <- players %>%
  calculate_summary_by(team) %>% 
  left_join(records) %>%
  select(-year)

team_position_summary <- players %>%
  calculate_summary_by(team, position) %>% 
  left_join(records) %>%
  select(-year)

position_summary
team_summary
team_position_summary
```

## Position Height, Weight Analyses

```{r}
position_summary %>%
  arrange(height) %>%
  ggplot(aes(x=height, y=reorder(position, height), label=round(height, 1))) +
    geom_bar(stat="identity", fill="navy") +
    geom_text(nudge_x=2.5) +
    theme_minimal() +
    ggtitle("Average Height by Position Group") +
    xlab("Average Height (inches)") +
    ylab("Position Group") +
    theme(plot.title=element_text(face = "bold"))

position_summary %>%
  arrange(weight) %>%
  ggplot(aes(x=weight, y=reorder(position, weight), label=round(weight, 1))) +
    geom_bar(stat="identity", fill="navy") +
    geom_text(nudge_x=max(position_summary$weight) * 0.04) +
    theme_minimal() +
    ggtitle("Average Weight by Position Group") +
    xlab("Average Weight (pounds)") +
    ylab("Position Group") +
    theme(plot.title=element_text(face = "bold"))

position_summary %>%
  arrange(bmi) %>%
  ggplot(aes(x=bmi, y=reorder(position, bmi), label=round(bmi, 1))) +
    geom_bar(stat="identity", fill="navy") +
    geom_text(nudge_x=max(position_summary$bmi) * 0.04) +
    theme_minimal() +
    ggtitle("Average BMI by Position Group") +
    xlab("Average Body Mass Index (BMI; kg/m^2)") +
    ylab("Position Group") +
    theme(plot.title=element_text(face = "bold"))
```

```{r}
height_ANOVA <- players %>%
  clean_position_data() %>% 
  aov(height ~ position, .)

summary(height_ANOVA)

weight_ANOVA <- players %>%
  clean_position_data() %>% 
  aov(weight ~ position, .)

summary(weight_ANOVA)

bmi_ANOVA <- players %>%
  clean_position_data() %>% 
  aov(bmi ~ position, .)

summary(bmi_ANOVA)
```

```{r}
TukeyHSD(height_ANOVA, conf.level=0.95,)$position %>%
  as.data.frame() %>%
  arrange(desc(`p adj`))

TukeyHSD(weight_ANOVA, conf.level=0.95)$position %>%
  as.data.frame() %>%
  arrange(desc(`p adj`))

TukeyHSD(bmi_ANOVA, conf.level=0.95)$position %>%
  as.data.frame() %>%
  arrange(desc(`p adj`))


# Formatting for Appendix:
# TukeyHSD(weight_ANOVA, conf.level=0.95)$position %>%
#   as.data.frame() %>%
#   arrange(desc(`p adj`)) %>%
#   select(`p adj`) %>%
#   filter(`p adj` > 0.05) %>%
#   select(-`p adj`)
```

## Binomial Regression

### Format Data for Regression

```{r}
position_heights_by_team <- team_position_summary %>%
  select(team, position, height) %>%
  pivot_wider(names_from = position, values_from = height, names_prefix = "height_")

position_weights_by_team <- team_position_summary %>%
  select(team, position, weight) %>%
  pivot_wider(names_from = position, values_from = weight, names_prefix = "weight_")

position_bmis_by_team <- team_position_summary %>%
  select(team, position, bmi) %>%
  pivot_wider(names_from = position, values_from = bmi, names_prefix = "bmi_")

team_position_summary_wide <- records %>%
  select(-year) %>%
  left_join(position_heights_by_team) %>%
  left_join(position_weights_by_team) %>% 
  left_join(position_bmis_by_team)

team_position_summary_wide
```

### Aside on Properly Performing Binomial Regression in R

Performing true _binomial_ regression in R (NOT logistic) is quite tricky and requires very careful use of the parameters to `glm`. The are two main approaches:

 1. Explicitly providing the number of binomial successes / failures in the response variable:
 
 Buried in the `glm` documentation:
 > For binomial and quasibinomial families the response can also be specified as a factor (when the first level denotes failure and all others success) or as a two-column matrix with the columns giving the numbers of successes and failures.
 
 2. Passing the proportion of successes along with the total number of trials as :
 
 Buried in the `glm` documentation:
 > For a binomial GLM prior `weights` are used to give the number of trials when the response is the proportion of successes

The two approaches are functionally equivalent, but can produce slightly different outputs due to rounding issues. You can see this in this cell:

```{r}
team_summary %>%
  glm(
    cbind(wins, losses) ~ height + weight + bmi,
    family = "binomial",
    data=.,
  ) %>%
  summary()

team_summary %>%
  glm(
    win_percentage ~ height + weight + bmi,
    weights = games,
    family = "binomial",
    data=.,
  ) %>%
  summary()
```

The 2 models in this cell show that when there is no complicated rounding, the two approaches above yield identical outputs:

```{r}
team_summary %>%
  filter(
    games == 12,
    wins != 7,
    wins != 5,
    wins != 4,
    wins != 2,
    wins != 1,
  ) %>%
  glm(
    cbind(wins, losses) ~ height + weight + bmi,
    family = "binomial",
    data=.,
  ) %>%
  summary()

team_summary %>%
  filter(
    games == 12,
    wins != 7,
    wins != 5,
    wins != 4,
    wins != 2,
    wins != 1,
  ) %>%
  glm(
    win_percentage ~ height + weight + bmi,
    weights = games,
    family = "binomial",
    data=.,
  ) %>%
  summary()
```

For maximum accuracy, I will prefer the explicit successes / failures approach so as to not have the function introduce small errors by rounding an already rounded proportion.

### Overall Height, Weight, BMI

```{r}
team_summary %>%
  glm(
    cbind(wins, losses) ~ height + weight + bmi,
    family = "binomial",
    data=.,
  ) %>%
  summary()
```

### Overall Height & Weight

```{r}
team_summary %>%
  glm(
    cbind(wins, losses) ~ height + weight,
    family = "binomial",
    data=.,
  ) %>%
  summary()
```

### Overall Univariate

```{r}
team_summary %>%
  glm(
    cbind(wins, losses) ~ height,
    family = "binomial",
    data=.,
  ) %>%
  summary()

team_summary %>%
  glm(
    cbind(wins, losses) ~ weight,
    family = "binomial",
    data=.,
  ) %>%
  summary()

team_summary %>%
  glm(
    cbind(wins, losses) ~ bmi,
    family = "binomial",
    data=.,
  ) %>%
  summary()
```

```{r}
log_odds_to_prob <- function(log_odds) {
  odds <- exp(log_odds)
  odds / (1 + odds)
}

win_prob_change_weight <- function(weight_change) {
  intercept <- -10.56900
  coeff <- 0.04653
  mean_weight <- mean(team_summary$weight)
  (
    log_odds_to_prob(intercept + coeff * (mean_weight + weight_change))
    - log_odds_to_prob(intercept + coeff * mean_weight)
  )
}

win_prob_change_weight(-10)
win_prob_change_weight(-1)
win_prob_change_weight(1)
win_prob_change_weight(10)
1/12
win_prob_change_weight(7.341706)
```


### All Heights, Weights, BMIs by Position Group

```{r}
team_position_summary_wide %>%
  glm(
    cbind(wins, losses) ~ . -team -games -win_percentage,
    family = "binomial",
    data=.,
  ) %>%
  summary()
```

### All BMIs by Position Group

```{r}
team_position_summary_wide %>%
  glm(
    cbind(wins, losses) ~
      bmi_LS +
      bmi_PK +
      bmi_P +
      bmi_WR +
      bmi_OL +
      bmi_LB +
      bmi_RB +
      bmi_QB +
      bmi_DB +
      bmi_DL +
      bmi_TE,
    family = "binomial",
    data=.,
  ) %>%
  summary()
```

### Maximizing Significant Features - Height & Weight

```{r}
best_model_height_weight <- team_position_summary_wide %>%
  glm(
    cbind(wins, losses) ~
      # height_LS +  # Exclude special teams
      # height_P +
      # height_PK +
      # height_LB +  # Least significant
      # height_DB +  # 2nd least significant
      # height_RB +  # Etc.
      # height_WR +
      # height_OL +
      # height_DL +
      # height_TE +
      height_QB +
      # weight_LS +  # Exclude special teams
      # weight_PK +
      # weight_P +
      # weight_WR +  # Least significant
      # weight_OL +  # 2nd least significant
      # weight_LB +  # Etc.
      # weight_RB +
      # weight_QB +
      weight_DB +
      weight_DL +
      weight_TE,
    family = "binomial",
    data=.,
  )

best_model_height_weight %>%
  summary()
```

```{r}
win_prob_change_position_weight <- function(QB=0, DB=0, DL=0, TE=0) {
  baseline <- best_model_height_weight %>%
    predict(newdata=data_frame(
      height_QB=mean(team_position_summary_wide$height_QB, na.rm=TRUE),
      weight_DB=mean(team_position_summary_wide$weight_DB, na.rm=TRUE),
      weight_DL=mean(team_position_summary_wide$weight_DL, na.rm=TRUE),
      weight_TE=mean(team_position_summary_wide$weight_TE, na.rm=TRUE)
    ))
  
  new <- best_model_height_weight %>%
    predict(newdata=data_frame(
      height_QB=mean(team_position_summary_wide$height_QB, na.rm=TRUE) + QB,
      weight_DB=mean(team_position_summary_wide$weight_DB, na.rm=TRUE) + DB,
      weight_DL=mean(team_position_summary_wide$weight_DL, na.rm=TRUE) + DL,
      weight_TE=mean(team_position_summary_wide$weight_TE, na.rm=TRUE) + TE
    ))
  
  (
    log_odds_to_prob(new[[1]])
    - log_odds_to_prob(baseline[[1]])
  )
}

win_prob_change_position_weight(DB=-1)
win_prob_change_position_weight(DL=-1)
win_prob_change_position_weight(TE=-1)

win_prob_change_position_weight(DB=1)
win_prob_change_position_weight(DL=1)
win_prob_change_position_weight(TE=1)

win_prob_change_position_weight(DB=10)
win_prob_change_position_weight(DL=10)
win_prob_change_position_weight(TE=10)

win_prob_change_position_weight(DB=9.46)
win_prob_change_position_weight(DL=27.42)
win_prob_change_position_weight(TE=12.65)
```


### Maximizing Significant Features - BMI

```{r}
team_position_summary_wide %>%
  glm(
    cbind(wins, losses) ~
      # bmi_LS +  # Exclude special teams
      # bmi_PK +
      # bmi_P +
      # bmi_OL +
      # bmi_WR +
      # bmi_RB +
      # bmi_QB +
      # bmi_LB +
      bmi_DB +
      bmi_TE +
      bmi_DL,
    family = "binomial",
    data=.,
  ) %>%
  summary()
```

## Scatter Plots

```{r}

team_summary %>% 
  ggplot(aes(x=height, y=win_percentage)) +
    geom_point() +
    theme_minimal() +
    ggtitle("Average Team Height vs. Win Percentage") +
    xlab("Average Height (inches)") +
    ylab("Win Percentage") +
    scale_y_continuous(labels = scales::percent) +
    theme(plot.title=element_text(face = "bold"))

team_summary %>% 
  ggplot(aes(x=weight, y=win_percentage)) +
    geom_point() +
    theme_minimal() +
    ggtitle("Average Team Weight vs. Win Percentage") +
    xlab("Average Weight (pounds)") +
    ylab("Win Percentage") +
    scale_y_continuous(labels = scales::percent) +
    theme(plot.title=element_text(face = "bold"))

team_summary %>% 
  ggplot(aes(x=bmi, y=win_percentage)) +
    geom_point() +
    theme_minimal() +
    ggtitle("Average Team BMI vs. Win Percentage") +
    xlab("Average Body Mass Index (BMI; kg/m^2)") +
    ylab("Win Percentage") +
    scale_y_continuous(labels = scales::percent) +
    theme(plot.title=element_text(face = "bold"))

```

```{r}
team_position_summary %>%
  filter(position == "TE") %>%
  ggplot(aes(x=weight, y=win_percentage)) +
    geom_point() +
    theme_minimal() +
    ggtitle("Average TE Weight vs. Win Percentage") +
    xlab("Average Weight (pounds)") +
    ylab("Win Percentage") +
    scale_y_continuous(labels = scales::percent) +
    theme(plot.title=element_text(face = "bold"))

team_position_summary %>%
  filter(position == "DL") %>%
  ggplot(aes(x=weight, y=win_percentage)) +
    geom_point() +
    theme_minimal() +
    ggtitle("Average DL Weight vs. Win Percentage") +
    xlab("Average Weight (pounds)") +
    ylab("Win Percentage") +
    scale_y_continuous(labels = scales::percent) +
    theme(plot.title=element_text(face = "bold"))

team_position_summary %>%
  filter(position == "DB") %>%
  ggplot(aes(x=weight, y=win_percentage)) +
    geom_point() +
    theme_minimal() +
    ggtitle("Average DB Weight vs. Win Percentage") +
    xlab("Average Weight (pounds)") +
    ylab("Win Percentage") +
    scale_y_continuous(labels = scales::percent) +
    theme(plot.title=element_text(face = "bold"))

```

```{r}
team_position_summary %>%
  filter(position == "QB") %>%
  ggplot(aes(x=height, y=win_percentage)) +
    geom_point() +
    theme_minimal() +
    ggtitle("Average QB Height vs. Win Percentage") +
    xlab("Average Height (inches)") +
    ylab("Win Percentage") +
    scale_y_continuous(labels = scales::percent) +
    theme(plot.title=element_text(face = "bold"))

```

```{r}
team_position_summary %>%
  filter(position == "DL") %>%
  ggplot(aes(x=bmi, y=win_percentage)) +
    geom_point() +
    theme_minimal()

team_position_summary %>%
  filter(position == "TE") %>%
  ggplot(aes(x=bmi, y=win_percentage)) +
    geom_point() +
    theme_minimal()

team_position_summary %>%
  filter(position == "DB") %>%
  ggplot(aes(x=bmi, y=win_percentage)) +
    geom_point() +
    theme_minimal()

```

## Conditional Plots

```{r}
# Plot Helpers
plot_win_percent_by_quartile <- function(df, metric, metric_title) {
  df %>%
    ungroup() %>%
    mutate(quartile = ntile({{ metric }}, 4)) %>% 
    group_by(quartile) %>%
    summarize(
      win_percentage = mean(win_percentage),
      n = sum(n)
    ) %>%
    ggplot(aes(x=quartile, y=win_percentage, label=scales::percent(win_percentage, 1))) +
      geom_col() +
      geom_text(nudge_y=0.015) +
      theme_minimal() +
      ggtitle(glue("Expected Win Percentage Given {metric_title} Quartile")) +
      xlab(glue("Average {metric_title} Quartile")) +
      ylab("Win Percentage") +
      scale_y_continuous(labels = scales::percent) +
      theme(plot.title=element_text(face = "bold"))
}
```


```{r}
team_summary %>%
  plot_win_percent_by_quartile(height, "Height")

team_summary %>%
  plot_win_percent_by_quartile(weight, "Weight")

team_summary %>%
  plot_win_percent_by_quartile(bmi, "BMI")

```

```{r}
team_position_summary %>%
  filter(position == "QB") %>%
  plot_win_percent_by_quartile(height, "QB Height")

```

```{r}
team_position_summary %>%
  filter(position == "DL") %>%
  plot_win_percent_by_quartile(weight, "DL Weight")

team_position_summary %>%
  filter(position == "TE") %>%
  plot_win_percent_by_quartile(weight, "TE Weight")

team_position_summary %>%
  filter(position == "DB") %>%
  plot_win_percent_by_quartile(weight, "DB Weight")

```

```{r}
team_position_summary %>%
  filter(position == "DL") %>%
  plot_win_percent_by_quartile(bmi, "DL BMI")

team_position_summary %>%
  filter(position == "TE") %>%
  plot_win_percent_by_quartile(bmi, "TE BMI")

team_position_summary %>%
  filter(position == "DB") %>%
  plot_win_percent_by_quartile(bmi, "DB BMI")

```

